Efficient forwards-time simulation with ancestral recombination graphs

### 2017-11-06 forwards_paper.tex

Another critical assumption of the coalescent is that the sample size $n$ is much smaller
than the effective population size $N_e$. With the very large sample sizes becoming common
in human genetics, this is no longer a safe assumption. For example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
reference panels in genome wide association studies. While many results of the coalescent are surpisingly
robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
are distorted and coalescent simulations may be
misleading~\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
A hybrid approach, in which the recent history of a large sample of
individuals is simulated under a detailed forwards-in-time model and the deep history
under the coalescent, is an attractive prospect.
\jk{Not sure where this should go; perhaps in the discussion? We can make the point that
by keeping track of the genealogies forwards in time we can easily combine these with the
deep history simulated by msprime}

@@

%%%%%%
\subsection*{Tree sequences}
A tree sequence is an efficient encoding for a sequence of correlated trees such as
is produced by recombination. The encoding is efficient because branches that are
shared by adjacent trees are stored once, rather than repeatedely for each tree.
The topology of a tree sequence is defined via a set of \emph{nodes} and
\emph{edges}. A node in a tree sequence refers to a distinct ancestor and
corresponds to the vertices in the individual genealogies along the sequence.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with. In the example of Figure~\ref{fig:example_tree_sequence}
we have a total of five nodes. Nodes 0, 1 and 2 occur at time 0 and are our
samples. Nodes 3 and 4 represent the ancestors of these samples, and were born at
time 1.5 and 2.5 respectively.

@@

This algorithm records only the topological information resulting from the
forwards in time Wright-Fisher process, but it is straightforward to add
mutational information. This can be done in two different ways.
We can record mutations that occur during the simulation quite simply.
For example, in Algorithm~\algref{W} we would generate mutations after
we have recorded the edges joining the new node $n$ to its parents
$P_a$ and $P_b$ in step \algref{W7}.
For example, if we assume that infinite sites mutations
occur at rate $\mu$ per generation then there are $\mbox{Poisson}\left((r - \ell)(\tau_p
- \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
will occur on a discinct site $x$ drawn uniformly from $(\ell, r]$ and be over node $c$.
It is straightforward to record this information during the simulation, but it
is significantly simpler and more efficient to generate these mutations
after the simulation has completed. In the case of infinite sites mutations
there is no difference between generating these mutations during the forwards
in time simulation and after it has completed. We are using precisely the same
information.

\begin{figure}
    \begin{center}
        \includegraphics{wf-before}
        \includegraphics{wf-after}
    \end{center}
    \caption{\label{fig:wf-trees}
    }
\end{figure}

It is significantly more efficient to generate mutations as a separate process
after the topological simulation has completed because we will generate many
mutations that will be lost in the population.

\begin{enumerate}
\item Explain that we can input these trees into an mpsrime tree sequence and they will
work fine. Perhaps a diagram to show the kind of trees we get? (Multiroots, lots of unary nodes.
Can use the msprime SVG API to draw these?) We could refer ahead to figure 2
instead though.
\item However, there is huge amount of redundancy in these. The amount of information stored
grows linearly with time.

\end{enumerate}

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the method.
        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
        It is recorded, in forwards time,
        in 10 node records (one for each individual)
        and 14 edgeset records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$
        to ensure rootnedness of resulting tree.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red,
        which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

@@

The resulting tables encode everything we need to know --
in fact, they record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of a smaller sample of indidivuals.
Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
we call \emph{simplification}.
Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
adding node and edgeset records to the output only when coalescent events are reached.
This works exactly as in \msprime, allowing substantial re-use of algorithms;
the main difference being that parental choice, mutations, and recombination locations
are determined by the input tree sequence
rather than randomly generated.
An example is shown in Figure \ref{fig:method_diagram}.

@@

\plr{Something about putting mutations down on the tree sequence.}

@@

\plr{Quick overview of how to efficiently hook this up with other code.}

@@

\plr{Estimates of run-time complexity}
Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
in which there are $S$ selected loci and $L$ neutral loci.
We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
and an ``ancestry-tracking'' strategy like that we consider here.
To do this, we assume that each individual must carry along its entire genotype.
More advanced schemes are used in some simulators,
but these increase efficiency by utilizing redundancy introduced by shared ancestry,
which is effectively an intermediate scheme.
We omit the cost of computing a fitness function.

Both schemes must choose mates and recombination breakpoints,
and pass on selected genotypes.
The difference between the two comes from the tradeoff between
(a) passing on neutral genotypes, and
(b) recording and simplifying the tree sequence, and adding neutral genotypes afterwards.
(We assume here that selected genotypes are stored in the same way for both.)
Passing on neutral genotypes naively records $L$ items per individual each generation, discarding the previous generation.
\plr{do this better with numbers from nodes, edgesets}
Recording a tree sequence stores 2 parents and $\rho$ breakpoints on average each generation;
after $T$ generations this grows to $N \times T \times (2 + \rho)$ stored items.
However, after simplification a tree sequence for all $N$ individuals
only takes of order $N + \rho \log N$ records.
Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;

>>>>

Another critical assumption of the coalescent is that the sample size $n$ is much smaller
than the effective population size $N_e$. With the very large sample sizes becoming common
in human genetics, this is no longer a safe assumption. For example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
reference panels in genome wide association studies. While many results of the
coalescent are surprisingly
robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
are distorted and coalescent simulations may be
misleading~\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
A hybrid approach, in which the recent history of a large sample of
individuals is simulated under a detailed forwards-in-time model and the deep history
under the coalescent, is an attractive prospect.
\jk{Not sure where this should go; perhaps in the discussion? We can make the point that
by keeping track of the genealogies forwards in time we can easily combine these with the
deep history simulated by msprime}

@@

%%%%%%
\subsection*{Tree sequences}
A tree sequence is an efficient encoding for a sequence of correlated trees such as
is produced by recombination. The encoding is efficient because branches that are
shared by adjacent trees are stored once, rather than repeatedly for each tree.
The topology of a tree sequence is defined via a set of \emph{nodes} and
\emph{edges}. A node in a tree sequence refers to a distinct ancestor and
corresponds to the vertices in the individual genealogies along the sequence.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with. In the example of Figure~\ref{fig:example_tree_sequence}
we have a total of five nodes. Nodes 0, 1 and 2 occur at time 0 and are our
samples. Nodes 3 and 4 represent the ancestors of these samples, and were born at
time 1.5 and 2.5 respectively.

@@

This algorithm records only the topological information resulting from the
forwards in time Wright-Fisher process, but it is straightforward to add
mutational information. This can be done in two different ways.
We can record mutations that occur during the simulation quite simply.
For example, in Algorithm~\algref{W} we would generate mutations after
we have recorded the edges joining the new node $n$ to its parents
$P_a$ and $P_b$ in step \algref{W7}.
For example, if we assume that infinite sites mutations
occur at rate $\mu$ per generation then there are $\mbox{Poisson}\left((r - \ell)(\tau_p
- \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
will occur on a distinct site $x$ drawn uniformly from $(\ell, r]$ and be over node $c$.
It is straightforward to record this information during the simulation, but it
is significantly simpler and more efficient to generate these mutations
after the simulation has completed. In the case of infinite sites mutations
there is no difference between generating these mutations during the forwards
in time simulation and after it has completed. We are using precisely the same
information.

\begin{figure}
    \begin{center}
        \includegraphics{wf-before}
        \includegraphics{wf-after}
    \end{center}
    \caption{An example of a marginal genealogy from a Wright-Fisher simulation
    with $N=5$. On the left we have the original tree including all
    intermediate nodes and dead-ends. On the right we have the minimal tree
    relating all of the currently-alive individuals (27--31).
    \jk{This figure needs a little tidying up. We should combine the two SVGs
        into a single image with some (a) (b) annotations. Also the node labels
        are poorly aligned. The SVGs are generated by the sims/wf-figures.py.}
    \label{fig:wf-trees}
    }
\end{figure}

It is significantly more efficient to generate mutations as a separate process
after the topological simulation has completed because we will generate many
mutations that will be lost in the population. Figure~\ref{fig:wf-trees} shows
an example of a marginal genealogy produced by a forwards-time Wright-Fisher
process like Algorithm~\algref{W}. There are two different trees in this
figure: one shows all the edges output by the simulation and the other
is the minimal tree representing the ancestry of the currently alive
population. Clearly there is a great deal of redundancy in the topological
information output by the simulation.

There are two sources of redundancy here. The first type of redundancy arises
from nodes in tree that have only one child. In Algorithm~\algref{W} we do
not attempt to track coalescence events but simply record all parent-child
relationships in the history of the population. As such, my of these edges
will record the simple passing of genealogical information from parent to child
and only some small subset will correspond to coalesences within the marginal
trees. The second source of redundancy in the output of Algorithm~\algref{W}
is due to the fact that lineages die out: there will be a large number of
individuals in the simulation that leave no ancestors in the present day
population. Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
ancestors in the current population and so the entire path tracing back to
the root is redundant.

One of the reasons that it is more efficient to generate mutations on the
genealogies after we have completed the forwards in time simulation is
because we avoid the cost of generating mutations on these dead-end branches.
It is only with the benefit of hindsight that we can know which lineages end
up being ancestral to the sample, and we can avoid a substantial cost of
generating and maintaining mutational information on these evolutionary
dead-ends. Assuming mutations are rare, another  reason that it is more efficient
to place mutations on the trees afterwards is that we have far fewer edges
in the trees: because we have removed all the intermediate `unary' nodes, we
have fewer edges to generate mutations on.

There are many advantages to having minimal genealogies such as show in
Figure~\ref{fig:wf-trees}. Computing this minimal representation of the
edges output by a forward-time simulation is an instance of a more
general problem that we refer to as `tree sequence simplification'. We discuss
this problem and an efficient solution in the next subsection.

@@

Suppose we have a tree sequence consisting of a set of nodes and edges as
described earlier, and we have some subset of the nodes in this tree sequence
that we are interested in (our `samples').  We wish to reduce the input tree sequence
into the minimum representation of the topologies that include the specified
samples. The output tree sequence must have the following properties:
\begin{enumerate}
\item we must have the same marginal trees with respect to the samples as
the input;
\item within the marginal trees, all non-sample vertices must has at least
two children (i.e., unary tree vertices are removed);
\item any nodes and edges not reachable from the sampled nodes are removed;
\item there are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
c)$ and $(x, r, p, c)$ which can be represented with a single edge
$(\ell, r, p, c)$.
\end{enumerate}

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the method.
        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
        It is recorded, in forwards time,
        in 10 node records (one for each individual)
        and 14 edge records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red,
        which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

In the current context of running forwards in time simulations, the tables of
nodes and edges that we record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of the current day population. We can use the process of simplification here to
reduce this down to the minimal set of nodes and edges required to represent
the information that we are interested in. We can also use simplification if  we have some very
large tree sequence representing a large dataset and we wish to extract the
information relevant to a subset of the samples.

Roughly, simplification works by tracing ancestry from the samples backwards
through the recorded history, adding node and edge records to the output
only when coalescent events are reached. This works exactly as in \msprime,
allowing substantial re-use of algorithms; the main difference being that
parental choice, mutations, and recombination locations are determined by the
input tree sequence rather than randomly generated. An example is shown in
Figure \ref{fig:method_diagram}.

The approach that we take is based on Hudson's algorithm for simulating
the coalescent with
recombination~\citep{hudson1983properties,kelleher2016efficient}. The main
state of the algorithm is a set of ancestral segments; each segment $(\ell, r,
u)$ represents a tract of ancestral material in which the node $u$ is
associated with the genomic interval $[\ell, r)$. In the same manner
as Hudson's algorithm, we represent the set of extant lineages as a
set of individuals, where each ancestral individual is composed of a
linked list of these segments. The node $u$ in each ancestral sequence
is the node ID associated with this segment in the output tree sequence.
We maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
the input tree sequence.

\jk{I'm not sure this is the right level of detail to tackle this description
at. What's the goal of this algorithm description? Are we trying to
establish some properties of the algorithm or just give the reader a high-level
flavour of how it works??}

@@

@@

The \msprime\ Python API provides a powerful platform for working with
tree topology and mutation data. We refer to the part of \msprime\
that is dedicated to tree sequence input and output as the `Tables API',
as the API is organised around simple tables of data. There are four key
tables: nodes, edges, sites and mutations. Briefly, nodes and edges define
the topology of a tree sequence (as defined above) and the sites and mutations
define mutational processes on this topology.
Figure~\ref{fig:example_tree_sequence} gives an informal depiction of this
encoding in terms of these tables.

The tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program.
Following the current best-practises [citations: apache arrow, etc] data is stored
in a columnar format. There are many advantages to this approach, but the
principle advantage for our purposes is that it allows for very efficient
interchange of large amounts of numerical data. In principle, this enables
zero-copy semantics, where a data consumer can read the information directly
from the memory of a producer without incurring the overhead of a copy
[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
data from Python into the low-level C library used to manipulate
tree sequences.

The tables API provides basic input and output operations via the numpy
array interface, which provides a great deal of flexibility as well
as efficiently. This makes transferring data from various sources
such as HDF5 [citation], Dask, Zarr etc, straightforward. (For small
scale data and debugging purposes, a simple text format is also supported.)
Along with
these operations we also provide a function to sort a set of tables
to ensure that the records are in the form required to input
into an \msprime\ tree sequence object. The \texttt{simplify\_tables}
function implements tree sequence simplification, as described in the
previous subsection.

This interchange API is very efficient. [Describe a quick example where we generate
a many-gigabyte tree sequence using fwdpp, and the time required
to copy the node and edge data into the tables]. By using a simple numerical
encoding of tree topologies and contiguous arrays of data to store this
encoding, we can achieve data transfer rates that would be impossible under
any text-based approach while retaining excellent portability.

@@

\plr{Estimates of run-time complexity}
Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
in which there are $S$ selected loci and $L$ neutral loci using a similar
method to Algorithm~\algref{W}.
We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
and an ``ancestry-tracking'' strategy like that we consider here.
To do this, we assume that each individual must carry along its entire genotype.
More advanced schemes are used in some simulators,
but these increase efficiency by utilizing redundancy introduced by shared ancestry,
which is effectively an intermediate scheme.
We omit the cost of computing a fitness function.

Both schemes must choose mates and recombination breakpoints,
and pass on selected genotypes.
The difference between the two comes from the tradeoff between
(a) passing on neutral genotypes, and
(b) recording and simplifying the tree sequence, and adding neutral genotypes afterwards.
(We assume here that selected genotypes are stored in the same way for both.)
Passing on neutral genotypes naively records $L$ items per individual each
generation, discarding the previous generation, giving an overall complexity of
$O((L + S) NT)$. Tracking the $S$ selected loci and recording a tree sequence
using Algorithm~\algref{W} requires $O(SNT)$ time and space, since we
record at most two edges for each individual in each generation.
However, after simplification the tree sequence contains
$O(??)$ edges since []. The cost of generating $L$ neutral mutations
is then $O(L + ??)$. Assuming that $N$, $T$ and $L$ are large, this is
very much smaller than the straightforward $O((L + S) NT)$ time required
to simulate all mutations directly.

\jk{Not sure where $\rho$ comes in here now. Thinking about this, shouldn't we
be able to come up with some expression for the number of edges? Assuming
the very simple single breakpoint per generation of Alg W, what is the
expected number of breakpoints? This is a function of $N$ and $T$, right?
It's not clear to me how many breakpoints this ends up giving you in
the output tree sequence and how many edges this corresponds to.
It's probably worthwhile thinking about this a bit.}

% However, after simplification a tree sequence for all $N$ individuals
% only takes of order $N + \rho \log N$ records.
% Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;
